\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2016-12-14 document/XSGRB_sample.tex
Additional data and text. Plot for follow-up timing

\usepackage{natbib}

\bibliographystyle{mnras}
\bibpunct{(}{)}{;}{a}{}{,}    %% natbib cite format used by A&A and ApJ

@@

\title{\vspace{-0.5cm}The X-shooter GTO sample of GRB afterglow and host Galaxy spectra}
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}

@@

\author{
	J.~Selsing\inst{1}, 
	D.~Malesani\inst{1}, 
	T.~Kr\"{u}hler\inst{1}, 
	P.~Goldoni\inst{14}, 
	J.~P.~U. Fynbo\inst{1}, 
	A.~de~Ugarte~Postigo\inst{11}, 
	J.~Japelj\inst{20},
	P.~D'Avanzo,
	Z.~Cano,
	S.~Covino\inst{10}, 
	V.~D'Elia\inst{7, 12}, 
	H.~Flores,
	O.~E.~Hartoog\inst{6},
	J.~Hjorth\inst{1}, 
	P.~Jakobsson\inst{5}, 
	A.~Levan,
	A.~Melandri,
	S.~Piranomonte \inst{7},
	R.~S\'anchez-Ram\'irez\inst{11},
	S.~Schulze\inst{17, 18}, 
	N.~R.~Tanvir\inst{19},
	C.~Th{\"o}ne,
	S.~D.~Vergani\inst{7, 8},
	P.~M.~Vreeswijk\inst{3}, 
	D.~J.~Watson\inst{1},
	K.~Wiersema\inst{19},
	D.~Xu\inst{1}
	L.~Christensen\inst{1},
	A.~De~Cia\inst{3}, 
	L.~Kaper\inst{6}, 
	L.~A.~Antonelli,
	F.~Fiore,
	A.~Gomboc,
	P.~Groot,
	F.~Hammer,
	C.~Ledoux\inst{2}, 
	E.~Maiorano,
	B.~Milvang-Jensen\inst{1}, 
	E.~Palazzi,
	E.~Pian,
	J.~Schaye,
	G.~Tagliaferri\inst{7},
	R.~A.~M.~J.~Wijers\inst{6}
}

@@

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with 
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$
been identified (Noterdaeme et al.\ 2012, A\&A, 547, L1). Long GRB afterglow
spectra, by contrast, reveal such systems in the majority of cases (e.g., Jakobsson et
al.\ 2006, A\&A, 460, L13; Fynbo et al.\ 2009, ApJS, 185, 526).
Whereas DLAs towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due
to the atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing
redshifts (e.g., Rafelski et al.\ 2014, ApJL, 782, L29), GRBs allow us to see 
into the hearts of star-forming galaxies over
the full history of cosmic star formation from $z \approx 0$ to $z > 8$ (e.g.,
Tanvir et al.\ 2009, Nature, 461, 1254; Salvaterra et al.\ 2009, Nature, 461,
1258; Jakobsson et al.\ 2012, ApJ, 752, 62). With afterglow spectroscopy
(throughout the electromagnetic spectrum from X-rays to the sub-mm) we can hence
characterize the properties of star-forming galaxies over cosmic history in
terms of  redshifts, metallicities, molecular contents, ISM temperatures,
UV-flux densities, etc.. This is, however, only possible as long as there are
satellites in orbit that rapidly and accurately locate GRBs. The currently
operating \textit{Swift} satellite, launched in 2004 and still fully-functioning,
allows for very efficient follow-up observations of GRBs due to its
unprecedented rate, speed and precision of localisations.

%With GTO time on X-shooter we
%have started building a sample of afterglows with X-shooter spectroscopy.
%X-shooter is ideal for this with its high efficiency, wide wavelength coverage
%and spectral resolution sufficient for inferring the fundamental physical
%parameters just listed.

\smallskip
\smallskip

{\bf The case for a large sample of GRBs with X-shooter spectroscopy.}

There are
several unanswered but fundamental questions that must be addressed in order
to exploit the full potential of GRBs as cosmological probes.
%Among them: what is the true redshift distribution and corresponding luminosity
%function of long-duration GRBs?
More than 50\% of the \textit{Swift} bursts with measured redshift are at $z >
2$, and  5--10\% are expected to be above $z=5$ (Salvaterra et al. 2008, MNRAS, 380, L45; Salvaterra et al.\ 2012, ApJ, 749,
68; Jakobsson et al.\ 2012, ApJ, 752, 62; Perley et al.\ 2016, ApJ, 817, 7). 
A high redshift completeness is
crucial for our understanding of the link between the number density of GRBs
per unit redshift and the global star-formation history of the Universe, as
measured by other means (UV, FIR, sub-mm, see Robertson \& Ellis 2011, ApJ, 744,
95). The detection of GRBs at $z > 6$ shows that GRBs have become competitive as
a tool to identifying galaxies at the highest redshifts and unsurpassed in
providing detailed abundance information via absorption line spectroscopy
(Tanvir et al.\ 2012, ApJ, 754, 46, McGuire et al.\ 2016, ApJ submitted, 
arxiv:1512.07808).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Complete sample}	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

This sections should contain a plot og the follow-up delay and the acquisition image brightness at the time of observations.

@@

Due of the stochastic nature of the \lya-forest, the blue wing of the Lyman
alpha absorption line is randomly superposed with forest systems, making it
notoriously difficult to model. Similarly so, the red wing has got the ISM
signatures imprinted on it, especially strong absorption due to \SIii~which can
exhibit significant velocity structure. Along with instrumental effects, the
generative model for the data that we would use in a likelihood-based analysis
would be almost boundless in complexity, making formal fitting of the column
densities unfeasible.

@@

\begin{figure*}
	\centering
	\includegraphics[page=2, width=16cm]{figures/HI_measurements.pdf}
	\caption{Same as Fig. \ref{fig:HI1}}
	\label{fig:HI2}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=3, width=16cm]{figures/HI_measurements.pdf}
	\caption{Same as Fig. \ref{fig:HI1}}
	\label{fig:HI3}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=4, width=16cm]{figures/HI_measurements.pdf}
	\caption{Same as Fig. \ref{fig:HI1}}
	\label{fig:HI4}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=5, width=16cm]{figures/HI_measurements.pdf}
	\caption{Same as Fig. \ref{fig:HI1}}
	\label{fig:HI5}
\end{figure*}

@@

\bibliography{XSGRB_sample}

@@

\subsection{GRB090313}

\subsection{GRB100205A}
Observed 3 days after the \textit{Swift} trigger. No afterglow or host detected
in 10.8 ks. GRB likely located at high
redshift\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10399.gcn3}}. The spectrum
has not otherwise been published previously.

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

\usepackage{natbib}
\usepackage{caption}

\bibpunct{(}{)}{;}{a}{}{,}    %% natbib cite format used by A&A and ApJ

@@

\title{\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample}
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}

@@

\author{
	J.~Selsing\inst{1}, 
	J.~P.~U. Fynbo\inst{1}, 
	D.~Malesani\inst{1}, 
	T.~Kr\"{u}hler\inst{1}, 
	P.~Goldoni\inst{14}, 
	A.~de~Ugarte~Postigo\inst{11}, 
	J.~Japelj\inst{20},
	P.~D'Avanzo,
	Z.~Cano,
	S.~Covino\inst{10}, 
	V.~D'Elia\inst{7, 12}, 
	H.~Flores,
	O.~E.~Hartoog\inst{6},
	J.~Hjorth\inst{1}, 
	P.~Jakobsson\inst{5}, 
	A.~Levan,
	A.~Melandri,
	S.~Piranomonte \inst{7},
	R.~S\'anchez-Ram\'irez\inst{11},
	S.~Schulze\inst{17, 18}, 
	N.~R.~Tanvir\inst{19},
	C.~Th{\"o}ne,
	S.~D.~Vergani\inst{7, 8},
	P.~M.~Vreeswijk\inst{3}, 
	D.~J.~Watson\inst{1},
	K.~Wiersema\inst{19},
	D.~Xu\inst{1}
	L.~Christensen\inst{1},
	A.~De~Cia\inst{3}, 
	L.~Kaper\inst{6}, 
	L.~A.~Antonelli,
	F.~Fiore,
	A.~Gomboc,
	P.~Groot,
	F.~Hammer,
	C.~Ledoux\inst{2}, 
	E.~Maiorano,
	B.~Milvang-Jensen\inst{1}, 
	E.~Palazzi,
	E.~Pian,
	J.~Schaye,
	G.~Tagliaferri\inst{7},
	R.~A.~M.~J.~Wijers\inst{6}
}

@@

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep{Noterdaeme2012}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep{Jakobsson2006, Fynbo2009}. Whereas DLAs
towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
star-forming galaxies over the full history of cosmic star formation from $z
\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we can hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, etc.. This is, however,
only possible as long as there are satellites in orbit that rapidly and
accurately locate GRBs. The currently operating \textit{Swift} satellite,
launched in 2004 and still fully-functioning, allows for very efficient
follow-up observations of GRBs due to its unprecedented rate, speed and
precision of localisations.

There are several unanswered but fundamental questions that must be addressed in
order to exploit the full potential of GRBs as cosmological probes. More than
50\% of the \textit{Swift} bursts with measured redshift are at $z > 2$, and 
5--10\% are expected to be above $z=5$ \citep{Salvaterra2007,
	Salvaterra2012,Jakobsson2012, Perley2016}. A high redshift completeness is
crucial for our understanding of the link between the number density of GRBs per
unit redshift and the global star-formation history of the Universe, as measured
by other means \citep[UV, FIR, sub-mm, see][]{Robertson2012}. The detection of
GRBs at $z > 6$ shows that GRBs have become competitive as a tool to identifying
galaxies at the highest redshifts and unsurpassed in providing detailed
abundance information via absorption line spectroscopy \citep{Tanvir2012,
	McGuire2016}.

@@

In this section we describe the efficiency of the follow-up effort and the characteristics of the observed bursts. We also assess the degree to which the sample obtained is complete.

@@

A rapid follow-up is essential for the accurate designation of GRB the origin. The redshift determination of burst where the host is not easily

\begin{figure}
	\centerline{\includegraphics[width=8cm]{figures/timing.pdf}}
	\caption{Follow-up timing and afterglow magnitude at the start of observation. The points have been colored based on the redshift of the corresponding burst. }
	\label{fig:timing}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample completeness}	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 152 \todo{Get number} bursts has been triggered by \textit{Swift} in the period since the comissioning of VLT/X-shooter in \todo{Get date}. This constitutes the \textit{statistical sample}.

@@

Due of the stochastic nature of the \lya-forest, the blue wing of the Lyman
alpha absorption line is randomly superposed with forest systems, along with
strong absorption from \mnii~and \SIiii, making it notoriously difficult to
model. Additionally, Similarly so, the red wing has got the ISM signatures
imprinted on it, especially strong absorption due to \SIii, \sii and \nv~which
can exhibit significant velocity structure. Along with instrumental effects, the
generative model for the data that we would use in a likelihood-based analysis
would be almost boundless in complexity, making formal fitting of the column
densities outside the scope of this paper.

@@

\begin{figure*}
	\centering
	\includegraphics[page=2, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=3, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=4, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[page=5, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}

@@

\bibliographystyle{mnras}
\bibliography{XSGRB_sample}

@@

\subsection{GRB090313 (z = 3.373)}

\subsection{GRB090530 (z = 1.266)}

\subsection{GRB090809 (z = 2.737)}

\subsection{GRB090926 (z = 2.106)}

\subsection{GRB091018 (z = 0.971)}

\subsection{GRB091127 (z = 0.49)}

\subsection{GRB100205A  (z = na)}
Observed 3 days after the \textit{Swift} trigger. No afterglow or host detected
in 10.8 ks. GRB likely located at high
redshift\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10399.gcn3}}. The spectrum
has not otherwise been published previously.

@@

\subsection{GRB110128A (z=2.339)}

\subsection{GRB110407A (z=na)}

\subsection{GRB110709B (z=na)}

\subsection{GRB110715A (z=0.82)}

\subsection{GRB110721A (z=0.382)}

\subsection{GRB110808A (z=1.348)}

\subsection{GRB110818A (z=3.36)}

\subsection{GRB111005A (z=0.013)}

\subsection{GRB111008A  (z=4.989)}

\subsection{GRB111107A (z=2.893)}

\subsection{GRB111117A (z=1.3?)}

\subsection{GRB111123A  (z=3.151)}

\subsection{GRB111129A (z=na)}

\subsection{GRB111209A (z=0.677)}

\subsection{GRB111211A (z=0.478)}

\subsection{GRB111228A (z=0.716)}

@@

\subsection{GRB120923A (z = $\gtrsim8$?)}

@@

\subsection{GRB121201A (z = 3.385)}

\subsection{GRB121229A (z = 2.707)}

\subsection{GRB130131B (z = 2.539)}

@@

\subsection{GRB130418A (z = 1.218)}

\subsection{GRB130427A (z = 0.34)}

\subsection{GRB130427B (z = 2.78)}

\subsection{GRB130603B (z = 0.356)}

@@

\subsection{GRB130612A (z = 2.006)}

\subsection{GRB130615A (z = $~3$?)}

\subsection{GRB130701A (z = 1.155)}

\subsection{GRB131011A (z = 1.874)}

\subsection{GRB131030A (z = 1.296)}

\subsection{GRB131103A (z = 0.599)}

\subsection{GRB131105A (z = 1.686)}

\subsection{GRB131117A (z = 4.042)}

\subsection{GRB131231A (z =0.642)}

\subsection{GRB140114A (z = $\sim2.8$?)}

\subsection{GRB140213A (z = 1.208)}

\subsection{GRB140301A (z = 1.416)}

\subsection{GRB140311A (z = 4.95)}

\subsection{GRB140430A (z = 1.601)}

\subsection{GRB140506A  (z = 0.889)}

\subsection{GRB140515A (z = $\sim6.32$?)}

\subsection{GRB140614A (z = 4.233)}

\subsection{GRB140622A (z = 0.959)}

\subsection{GRB141028A (z = 2.332)}

\subsection{GRB141031A  (z = na)}

\subsection{GRB141109A (z = 2.993)}

\subsection{GRB150206A (z = 2.087)}

\subsection{GRB150301B (z = 1.517)}

\subsection{GRB150403A (z = 2.06)}

\subsection{GRB150423A (z = 1.394)}

\subsection{GRB150428A (z = na)}

\subsection{GRB150514A (z = 0.807)}

\subsection{GRB150518A (z = 0.256)}

\subsection{GRB150616A (z = 1.188)}

\subsection{GRB150727A (z = 0.313)}

\subsection{GRB150821A (z = 0.755)}

\subsection{GRB150831A (z = na)}

\subsection{GRB150910A (z = 1.359)}

\subsection{GRB150915A (z = 1.968)}

@@

\subsection{GRB151027B (z = 4.063)}

\subsection{GRB151029A (z = 1.423)}

\subsection{GRB151031A (z = 1.167)}

\subsection{GRB160117B (z = 0.87)}

@@

\subsection{GRB160228A (z = 1.64)}

\subsection{GRB160303A (z = na)}

\subsection{GRB160314A (z = 0.726)}

\subsection{GRB160410A (z = 1.717)}

\subsection{GRB160425A (z = 0.555)}

\subsection{GRB160625B (z = 1.406)}

@@

\subsection{GRB161001A (z = 0.891?)}

@@

\subsection{GRB161023A (z = 2.710)}

\subsection{GRB161117A (z = 1.549)}
